---
title: "yc4384_P8122_HW3"
author: "yc4384_Yangyang_Chen"
date: "`r Sys.Date()`"
output: 
  pdf_document:
      latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Data Preparation for Mice Study
```{r}
set.seed(124)
n_mice = 16  
p_obesity = 1/5  
obesity = rbinom(n_mice, 1, p_obesity)
theta_0 = 1/2  
theta_1 = -1/5  
p_light = theta_0 + theta_1 * obesity
light_exp = rbinom(n_mice, 1, p_light)  
beta_0 = 110  
beta_1 = 20  
beta_2 = -5 
sigma_Y = 1  
mu_glucose = beta_0 + beta_1 * obesity + beta_2 * light_exp  
glucose = rnorm(n_mice, mu_glucose, sigma_Y) 

mice_data = data.frame(
  mouse_id = 1:n_mice,
  obesity = obesity,
  light_exposure = light_exp,
  glucose = glucose
)

mice_data |> head()
```

## Problem 1: Interpretation of parameters:

$p$: Probability that the baseline covariate C (obesity) equals 1, indicating the mouse is obese (Bernoulli distribution).

$\theta_0$: Probability of assigning mice to light exposure when not obese $(C = 0)$.

$\theta_1$: Probability of assigning obese mice to light exposure is reduced by 1/5 compared to non-obese mice.

$\beta_0$: Average baseline glucose level when the mouse is not obese and unexposed to light.

$\beta_1$:  A unit increase in obesity results in an average increase of 20 units in glucose levels, holding other factors constant.

$\beta_2$: Exposure to light reduces glucose levels by 5 units on average, holding other factors constant.


## Problem 2: Marginal and Conditional PACE

The Marginal PACE is:

$$
E[Y_1] - E[Y_0] = \sum E[Y | A = 1, C = c] P(C = c) - \sum E[Y | A = 0, C = c] P(C = c)
$$

The Conditional PACE is:

$$
E[Y_1 | C = c] - E[Y_0 | C = c] = \beta_2 = -5
$$

I. Under the following assumptions Marginal PACE can be identified:

1) Consistency: 
$$
   Y = Y_1 \text{ if treated, and } Y = Y_0 \text{ if not treated}.
$$

2) Exchangeability: 
$$
   Y_1, Y_0 \perp A | X.
$$

3) Positivity: 
$$
   0 < P(A = 1 | X) < 1.
$$

4) SUTVA: No interference between units, and the treatment is well-defined for all units.

II. Under the following assumptions Conditional PACE can be identified:

1) Conditional Consistency:
$$
   Y = Y_1 \text{ if } A = 1, \quad Y = Y_0 \text{ if } A = 0, \quad \text{given } X.
$$

2) Conditional Exchangeability:
$$
   Y_1, Y_0 \perp A | X = x.
$$
   
3) Conditional Positivity:
$$
   0 < P(A = 1 | X = x) < 1.
$$

4) SUTVA: No interference between units, and the treatment is well-defined for all units.

## Problem 3: G-formula for Randomized and Observational Studies

g-formula for the randomized study:  $\sum_c E[Y | A = 1, C = c] Pr(C = c) - \sum_c E[Y | A = 0, C = c] Pr(C = c)=E[Y|A=1]-E[Y|A=0]$  

g-formula for the observational study: $\sum_c E[Y |A=1, C = c] Pr(C = c)-\sum_c E[Y |A=0, C = c] Pr(C = c)$ under NUCA.  

Randomization ensures unconfoundedness, meaning that the distribution of C is independent of  A  (the treatment assignment). As a result, the g-formula simplifies to  $E[Y|A=1] - E[Y|A=0]$ , as no adjustment for confounders is necessary. However, in an observational study,  A  is influenced by  C , meaning that the relationship between the treatment and outcome is confounded. In such cases, the g-formula cannot be simplified without appropriately accounting for the confounders. 

## Problem4: Estimate and confidence interval of $E[Y|A=1]-E[Y|A=0]$

The 95% confidence interval can be calculated by $E[Y|A=1]-E[Y|A=0] \pm z \sqrt{\frac{S_1^2}{N_1} + \frac{S_0^2}{N_0}}$, where $z$ is the critical value from the standard normal distribution, $S_1^2$ and $S_0^2$ are the variances of the potential outcomes in $A=1$ and $A=0$ groups, and $N_1$ and $N_0$ are the sizes of $A=1$ and $A=0$ groups:
```{r}
# Split Y into treatment and control groups
Y_1 = glucose[light_exp == 1]
Y_0 = glucose[light_exp == 0]

# Use pipeline to compute mean glucose levels and causal effect estimate
mean_Y_1 = Y_1 |> mean()
mean_Y_0 = Y_0 |> mean()

# Causal effect estimate
diff_estimate = mean_Y_1 - mean_Y_0

# Critical value for 95% CI
z_crit = qnorm(1 - 0.05 / 2)

# Variances and sample sizes
var_Y_1 = Y_1 |> var()
var_Y_0 = Y_0 |> var()
n_1 = Y_1 |> length()
n_0 = Y_0 |> length()

# Standard error
se_diff = (var_Y_1/n_1 + var_Y_0/n_0) |> sqrt()

# Confidence interval
ci_lower = diff_estimate - z_crit * se_diff
ci_upper = diff_estimate + z_crit * se_diff
```
Therefore,  the estimate of causal effect is $E[Y|A=1]-E[Y|A=0]=$ `r round(diff_estimate,2)`. The 95% confidence interval is (`r list(round(ci_lower, 2), round(ci_upper, 2))`), which excludes zero, indicating a causal effect between the bright and dark light groups in obesity. Therefore, the result is statistically significant at the 95% confidence level.

Additionally, the point estimate suggests that, on average, the outcome $Y$ is `r round(-diff_estimate,2)` mg/dL lower in bright light group compared to dark light group.

## 5. Estimate and confidence interval of $E[Y_1]-E[Y_0]$
The difference $E[Y_1]- E[Y_0]$  is calculated using the g-formula from Problem 3:
$E[Y_1]- E[Y_0]=\sum_c E[Y |A=1, C = c] Pr(C = c)-\sum_c E[Y |A=0, C = c] Pr(C = c)$

```{r}
p_C0 = obesity |> mean()
p_C1 = 1 - p_C0

# Mean outcomes for different combinations of A and C
mean_Y_0_0 = glucose[light_exp == 0 & obesity == 0] |> mean()
mean_Y_0_1 = glucose[light_exp == 0 & obesity == 1] |> mean()
mean_Y_1_0 = glucose[light_exp == 1 & obesity == 0] |> mean()
mean_Y_1_1 = glucose[light_exp == 1 & obesity == 1] |> mean()
```

Upon noticing that no obese mice were assigned to the bright light group, this violates the probabilistic assumption. Therefore, we will exclude the obese mice and consider only the non-obese group.

```{r}
# g-formula estimates
E_Y1 = p_C0 * mean_Y_1_0
E_Y0 = p_C0 * mean_Y_0_0

# Difference E[Y1] - E[Y0]
diff = E_Y1 - E_Y0
```

Therefore, $E[Y_1]- E[Y_0]=$ `r round(diff, 2)`

```{r}
# Critical value for 95% CI
z = qnorm(1 - 0.05 / 2)

# Sample sizes for A=1 and A=0, C=0
n_1_0 = sum(light_exp == 1 & obesity == 0)
n_0_0 = sum(light_exp == 0 & obesity == 0)

# Variances for A=1 and A=0, C=0
var_Y_1_0 = glucose[light_exp == 1 & obesity == 0] |> var()
var_Y_0_0 = glucose[light_exp == 0 & obesity == 0] |> var()

# Standard error
se_diff = (var_Y_1_0 / n_1_0 + var_Y_0_0 / n_0_0) |> sqrt()

# Confidence interval
ci_lower = diff - z * se_diff
ci_upper = diff + z * se_diff
```

The 95% confidence interval is (`r list(round(ci_lower, 2), round(ci_upper, 2))`).  

The point estimate indicates that, on average, the outcome $Y$ is `r round(-diff,2)` mg/dL  higher in the bright light group compared to the dark light group among non-obese mice in this observational study. However, due to the violation of probabilistic assumptions, we cannot infer the causal effect for obese mice. Since the 95% confidence interval includes zero, the point estimate is not statistically significant at the 95% confidence level.

The difference between the point estimates in Problems (4) and (5) arises because Problem (5) accounts for the adjustment of confounding bias.

# 6. Assumptions of estimate $E[Y_1]−E[Y_0]$ using linear regression with continuous covariates
In addition to consistency, SUTVA, exchangeability, positivity, and NUCA assumption, we need to assume the following:  
- No measurement error: All variables need to be correctly measured.
- No model misspecification: All models need to be correctly specified. In the linear regression, the relationship between $Y$, $A$, and covariates should be adequately approximated by a linear model.